\section{Integration by Parts}
\label{sec:integration-by-parts}
% Section 3.5
Integration By Parts
Integration by parts is an integration method which enables us to find antiderivatives of some new functions such as ln(x) as well as antiderivatives of products of functions such as x2ln(x) and xex.

If the function we're trying to integrate can be written as a product of two functions, u, and dv, then integration by parts lets us trade out a complicated integral for hopefully simpler one.

Integration by Parts Formula
∫udv=uv-∫vdu
For definite integrals:
∫baudv=uv]ba-∫bavdu
\begin{example}
Integrate ∫xexdx.

\begin{solution}
To use the Integration by Parts method, we break apart the product into two parts:
u=xanddv=exdx.
We now calculate du, the derivative of u, and v, the integral of dv:
du=(ddxx)dxandv=∫exdx=ex.
Using the Integration by Parts formula,
∫xexdx=uv-∫vdu=xex-∫exdx.
Notice the remaining integral is simpler that the original, and one which we can easily evaluate:
xex-∫exdx=xex-ex+C.
\end{solution}\end{example}

In the last example we could have chosen either x or ex as our u, but had we chosen u=ex, the second integral would have become messier, rather than simpler.

Rule of Thumb
When selecting the u for Integration by Parts, select a logarithmic expression if one is present. If not, select an algebraic expression (like x or dx).

(There is a larger decision tree that can be written down for choosing u and dv, but since we're not looking at any trigonometric functions in this course the rule above is sufficient for the functions we're integrating.)

\begin{example}
Integrate ∫146x2ln(x)dx.

\begin{solution}
Since this contains a logarithmic expression, we'll use it for our u:
u=ln(x)anddv=6x2dx
We now calculate du and v:
du=1xdxandv=∫6x2dx=6x33=2x3
Using the By Parts formula:
∫416x2ln(x)dx=2x3ln(x)]41-∫416x21xdx
We can simplify the expression in the integral on the right:
∫416x2ln(x)dx=2x3ln(x)]41-∫416xdx
The remaining integral is a basic one we can now evaluate:
∫416x2ln(x)dx=2x3ln(x)]41-3x2]41
Finally, we can evaluate the expressions:
∫416x2ln(x)dx==\approx   ((2⋅43ln(4))-(2⋅13ln(1)))-((3⋅42)-(3⋅12))128ln(4)-45132.446
\end{solution}\end{example}
